---
title: "Power analysis walkthrough"
output: html_document
date: "2025-05-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### running the function

This is a brief walkthrough of how to run the functions.

```{r source_scripts}

source("../scripts/functions/run_power_analysis.R")
source("../scripts/functions/power_subfunctions.R")

```

The steps behind the function:

1. Reads or processes the input datasets.

2. Filters rows (if sample_column is provided).

3. Extracts model parameters from real data using `get_model_pars`.

4. Creates a template data frame with simulated temporal and site structure using .

5. Splits site sampling across datasets based on data_proportions.

6. Simulates response data under the assumed model and effect sizes.

7. Fits GLMMs to each simulation and calculates how often the true fixed effects are detected.

8. Returns (and optionally saves) a summary data frame of the power analysis results.


### Parameter descriptions for `run_power_analysis`

Below are the explanations for each of the parameters that can be set for the `run_power_analysis` function.

`datasets`  
Accepts:
- A semicolon-separated character string of file paths (e.g., "data1.csv;data2.csv")
- A list of data frames
- A single data frame  
These are the source datasets used to extract model parameters and generate simulations.

`sample_column`  
(Optional) A character string naming a column used to subset the datasets. Should be a a column of `c(0,1)`, where 1s are the rows to keep. For example, if only rows where `sample_column == 1` should be used, specify it here.

`site_column`  
A character string naming the column that identifies unique sampling sites.

`model_pars`  
A character vector or semicolon-separated string listing all variables (covariates) to include in the model.

`fixed_effect`  
A character vector or semicolon-separated string specifying which of the model parameters should be treated as *fixed effects*. Must be a subset of `model_pars`. One of `fixed_effect` or `random_effect` is required.

`random_effect`  
Same as above, but for *random effects*. Also must be a subset of `model_pars`.  
One of `fixed_effect` or `random_effect` is required.

`yearly_samfreq`  
`Integer` Number of times each site is sampled per year (e.g., 12 for monthly, 1 for annual).

`yearly_samfreq_column`  
`Character` Name of the column that reflects the sampling frequency (default is "days").

`response_var`  
`Character` The name of the response variable in the dataset.

`effect.size`  
A numeric vector (or semicolon-separated string) representing the true effect sizes for the fixed effects. Must match the length and order of `fixed_effect`.

`nsim`  
`Integer` Number of simulations to run.

`samfreq`  
`Integer` Total number of samples per site across the entire study period.

`nosite.yr`  
`Integer` Number of unique sites sampled each year.

`noyear`  
`Integer` Number of years in the study.

`data_proportions`  
(Optional) A numeric vector (or semicolon-separated string) indicating what proportion of sites to simulate from each input dataset. Should sum to 1 and be the same length as the number of datasets.

`save_loc`  
(Optional) `Character` Directory where the output CSV should be saved. Creates the directory recursively if the location doesn't exist. If `NULL`, the results are not saved to disk.


#### examples

Here are some examples of how to run the power analysis function. We will only run the code for `nsim = 10` to reduce run times. The code below runs the power analysis function for one dataset and one fixed effect - `year`.

```{r}

pwr <- run_power_analysis(
  datasets = "../data/processed/dip_data_mn_5yrperc_change.csv", 
  sample_column = "sampled_20",
  site_column = "sampling_point",
  model_pars = c("month", "year", "sampling_point"), 
  random_effect = c("month", "sampling_point"), 
  yearly_samfreq = 12,
  yearly_samfreq_column = "month",
  response_var = "water_level",
  effect.size = 0.1,
  nsim = 10,
  samfreq = 1,
  nosite.yr = 100,
  noyear = 5,
  data_proportions = NULL,
  save_loc = NULL)
pwr


```

The code below runs the power analysis function for two datasets and one fixed effect - `year`.

```{r}

pwr <- run_power_analysis(
  datasets = paste("data/processed/dip_data_mn_5yrperc_change.csv", 
                   "data/processed/site_data_mn_5yrperc_change.csv", sep = ";"), 
  sample_column = "sampled_20",
  site_column = "sampling_point",
  model_pars = c("month", "year", "sampling_point"), 
  random_effect = c("month", "sampling_point"), 
  yearly_samfreq = 12,
  yearly_samfreq_column = "month",
  response_var = "perc_chg_ind",
  effect.size = 0.05,
  nsim = 10,
  samfreq = 1,
  nosite.yr = 100, 
  noyear = 5,
  data_proportions = c(0.8, 0.2), 
  save_loc = NULL)
pwr

```

Below we run the power analysis function for two datasets and two fixed effects - `month` and `year`. Each fixed effect also has a different `effect.size` over time `0.01` and `0.05`.

```{r}

pwr <- run_power_analysis(
  datasets = paste("data/processed/dip_data_mn_5yrperc_change.csv", 
                   "data/processed/site_data_mn_5yrperc_change.csv", sep = ";"), 
  sample_column = "sampled_20",
  site_column = "sampling_point",
  model_pars = c("month", "year", "sampling_point"), 
  random_effect = c("sampling_point"), 
  yearly_samfreq = 12,
  yearly_samfreq_column = "month",
  response_var = "perc_chg_ind",
  effect.size = c(0.01, 0.05),
  nsim = 10,
  samfreq = 1,
  nosite.yr = 100, 
  noyear = 5,
  data_proportions = c(0.8, 0.2), 
  save_loc = NULL)
pwr

```

